\documentclass{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{listings}
\usepackage{exercise}

\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
framexleftmargin=15pt,
basicstyle = \ttfamily,
columns=fullflexible
}

\title{Basic Building Blocks: Exercises}

\begin{document}
\maketitle

\section{Variables}

\begin{Exercise}

\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_var_1}, mathescape]
var x = 1
var y = x + 3
var z = y $\times$ 5
var w
\end{lstlisting}

\begin{enumerate}
\item What is the value of \lstinline{y} after Line 2 executes (or does it not exist or is this code invalid)?
\item What is the value of \lstinline{z} after Line 2 executes (or does it not exist or is this code invalid)?
\item What is the value of \lstinline{z} after Line 3 executes (or does it not exist or is this code invalid)?
\item What is the value of \lstinline{w} after Line 4 executes (or does it not exist or is this code invalid)?
\end{enumerate}

\end{Exercise}

\begin{Exercise}

\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_var_2}]
var x = x + 5
var y = 3
\end{lstlisting}

\begin{enumerate}
\item What is the value of \lstinline{x} after Line 2 executes (or does it not exist or is this code invalid)?
\end{enumerate}

\end{Exercise}

\begin{Exercise}

\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_var_3},mathescape]
var x = 2
var y = x $\times$ 2
var z = y $\times$ 2
x = z $\times$ 4
y = x / 2
\end{lstlisting}

\begin{enumerate}
\item What is the value of \lstinline{x} after Line 2 executes (or does it not exist or is this code invalid)?
\end{enumerate}

\begin{enumerate}
\item What is the value of \lstinline{x} after Line 3 executes (or does it not exist or is this code invalid)?
\item What is the value of \lstinline{y} after Line 5 executes (or does it not exist or is this code invalid)?
\item What is the value of \lstinline{z} after Line 2 executes (or does it not exist or is this code invalid)?
\end{enumerate}

\end{Exercise}

\section{Conditionals}

\begin{Exercise}
\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_conditionals_1},mathescape]
var x = 24
var y
IF x == 24 {
    y = 26
}

var z
IF x % 2 == 0 {
    z = 5
} ELSE {
    z = 10
}

IF x % 4 == 0 {
    y = 12
}
\end{lstlisting}

\begin{enumerate}
\item What is the value of \lstinline{x} after the program finishes running?
\item What is the value of \lstinline{y} after the program finishes running?
\item What is the value of \lstinline{z} after the program finishes running?
\end{enumerate}
\end{Exercise}

\begin{Exercise}

Write a program that assigns values to the variables \lstinline{y} and \lstinline{z} depending on the value of \lstinline{x} (choose your own initial value).
If \lstinline{x} is divisble by 7 then set \lstinline{y} to be $0$ and \lstinline{z} to be $x - 5$.
Otherwise, set \lstinline{y} to be $x + 1$ and \lstinline{z} to be $x - 1$.

\end{Exercise}

\section{Loops}

\begin{Exercise}
\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_loop_1},mathescape]
var x = 2
var i = 0
LOOP i $<=$ 7 {
    x = x + 5
}
\end{lstlisting}

\begin{enumerate}
\item How many times will the loop run?
\item What is the value of \lstinline{x} after the loop finishes running $3$ times?
\end{enumerate}
\end{Exercise}

\begin{Exercise}
\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_loop_2},mathescape]
var x = 1
var i = 9
LOOP i $>=$ 0 {
    x = x * 2
    i = i - 3
}
z = x / 2
\end{lstlisting}

\begin{enumerate}
\item How many times will the loop run?
\item What is the value of \lstinline{x} after the loop finishes running?
\item What is the value of \lstinline{z} after line 7 runs?
\end{enumerate}
\end{Exercise}

\begin{Exercise}

Write a program that will
\begin{enumerate}
\item Set the value of $y$ to be $7$.
\item Increment the value of $x$ (with an initial value of your choosing) $y$ number of times.
\end{enumerate}

\end{Exercise}

\begin{Exercise}

Write a program that will count the number of odd numbers between $1$ and $5258$ (inclusive).

\end{Exercise}

\section{Functions}

\begin{Exercise}
\begin{lstlisting}[caption={Pseudocode.}, label={code:exercise_fn_1},mathescape]
FUNCTION fn(var x) {
    IF x % 2 == 0 {
        return 1
    }
    return 5
}
var w = fn(20)
\end{lstlisting}

What is the value of \lstinline{w} after the program finishes running?

\end{Exercise}

\begin{Exercise}

Write a function signature for a function named \lstinline{callme} that takes in $2$ inputs.
Call the input variables whatever you want.

\end{Exercise}

\begin{Exercise}

Write a function that takes in three arguments: \lstinline{a}, \lstinline{b}, and \lstinline{x}.
The function should check all the numbers between \lstinline{a} and \lstinline{b} (exclusive) and return the numbers that are divisble by \lstinline{x}.

\end{Exercise}


\end{document}
